---
title: "Practical: extensions, spatial/temporal models and mixed responses"
subtitle: "Physalia workshop on GLLVMs"
author: "Bert van der Veen"
output: html_document
---

# Background

This practical covers two extensions to the basic GLLVM framework:

1. **Spatial and temporal autocorrelation**: accommodating non-independence of sampling units via structured covariance in the latent variables or row effects
2. **Mixed response types**: fitting models where different species have different response distributions

The key `gllvm` arguments for spatial/temporal structure are:

- `lvCor`: specifies correlation structure for the latent variables
- `row.eff`: specifies correlation structure for the row effects
- `distLV`: matrix of distances between sites (for spatial models)
- `corAR1`, `corExp`, `corMatern`, `corCS`: available correlation structures

Today's lecture demonstrated these ideas on small synthetic examples, plus a real case study: the Garchinger Heide calcareous grassland data (Bauer & Albrecht 2020), combining ordinal and zero-inflated count families with a spatial correlation structure. That dataset is not in the shared `data` folder, so here we work with wadden instead. If you want a dataset with real (rather than simulated) spatial or temporal replication, the <tt>gllvm</tt> package ships two:

1. wadden (abundance; simulated coordinates for the spatial part, real seasons for the temporal part; used below)
2. Skabbholmen (counts, `data(Skabbholmen)`; real repeated transects surveyed over several years, a genuine temporal succession dataset)
3. kelpforest (counts, `data(kelpforest)`; real site/year/transect structure, suitable for both the spatial and temporal parts)

# Part I: Temporal autocorrelation

We use a subset of the wadden sea data which has a temporal structure (seasons).

```{r setup, message=FALSE}
library(gllvm)
TMB::openmp(parallel::detectCores()-1, autopar = TRUE, DLL = "gllvm")

Y <- read.csv("../../data/waddenY2.csv")[, -c(1:2)]
Y <- Y[, colSums(ifelse(Y == 0, 0, 1)) > 2]
X <- read.csv("../../data/waddenX.csv")
X <- X[, !apply(X, 2, anyNA)]
X[, unlist(lapply(X, is.numeric))] <- scale(X[, unlist(lapply(X, is.numeric))])
```

Fit a baseline model (IID latent variables) and compare to one with AR(1) structure across seasons:

```{r ar1, cache=TRUE, message=FALSE, warning=FALSE}
# Baseline: independent latent variables
model_iid <- gllvm(Y, num.lv = 2, family = "tweedie", Power = NULL,
                   disp.formula = rep(1, ncol(Y)), n.init = 3)

# AR(1) structure across seasons in the row effect
model_ar1 <- gllvm(Y, num.lv = 2, family = "tweedie", Power = NULL,
                   disp.formula = rep(1, ncol(Y)),
                   studyDesign = X[, "season", drop = FALSE],
                   row.eff = ~corAR1(1|season), n.init = 3)

AIC(model_iid, model_ar1)
```

Inspect the estimated AR(1) correlation:

```{r ar1_param}
model_ar1$params$sigma  # the ".rho"-suffixed entry is the temporal correlation between seasons
```

## Tasks I

1. Fit both models and compare AIC. Does adding temporal structure improve fit?
2. Interpret the AR(1) correlation parameter. Is community composition persistent across seasons?
3. Try `lvCor = ~corAR1(1|season)` instead of `row.eff`. What is the difference in interpretation?
4. Visualize the ordination from both models. Does temporal structure change the ordination?

# Part II: Spatial models

If you have site coordinates, you can fit a model with spatial correlation. Here we simulate coordinates for illustration; in your own data, use real coordinates.

```{r spatial_setup}
# Simulate spatial coordinates for the wadden sites
set.seed(42)
n_sites <- nrow(Y)
coords <- data.frame(
  lon = runif(n_sites, 0, 100),
  lat = runif(n_sites, 0, 100)
)
dist_mat <- as.matrix(dist(coords))
```

```{r spatial, cache=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
model_spatial <- gllvm(Y, num.lv = 2, family = "tweedie", Power = NULL,
                       disp.formula = rep(1, ncol(Y)),
                       lvCor = ~corExp(1|site),
                       studyDesign = data.frame(site = factor(1:nrow(Y))),
                       distLV = dist_mat,
                       Lambda.struc = "UNN", NN = 5,
                       n.init = 1)
```

Note: spatial models are computationally demanding. With real data, ensure you have sufficient memory and consider sparse approximations (`Lambda.struc = "bdNN"` or `"UNN"` with `NN` set to a small number of neighbours).

## Tasks II

1. Read through the code for fitting a spatial model. What does `distLV` do?
2. What does the `NN` argument control? What are the trade-offs of using more vs. fewer neighbours?
3. How would you use `VP()` to see how much variation the spatial structure explains?
4. If you have your own spatial dataset, try fitting a `corExp` model to it.

# Part III: Mixed response types

Different species in a single dataset can call for different response distributions — for example, rare species recorded as presence-absence and common species recorded as counts. As covered in the lecture, `gllvm` supports this by passing a **character vector** to `family`, with one entry per species column (matching column order), rather than a single distribution for the whole community.

To illustrate on data we already have, we will treat the first 10 species in `Y` as presence-absence (thresholding to 0/1) and leave the rest as counts:

```{r mixed, cache=TRUE, message=FALSE, warning=FALSE}
Y_mixed <- Y
Y_mixed[, 1:10] <- ifelse(Y_mixed[, 1:10] > 0, 1, 0)

family_vec <- c(rep("binomial", 10), rep("negative.binomial", ncol(Y_mixed) - 10))

model_mixed <- gllvm(Y_mixed, num.lv = 2, family = family_vec, n.init = 3, sd.errors = FALSE)
```

## Tasks III

1. Compare `model_mixed` to a model that (incorrectly) treats all species in `Y_mixed` as counts (`family = "negative.binomial"`). What goes wrong for the thresholded species?
2. Think about your own data: do different species call for different distributions? Build a `family` vector for it, matching column order.
3. Check `?gllvm` for the full list of families that can be mixed together in one model.

# Part IV: Own data

Use the remaining time to apply what you have learned to your own dataset, or to explore further aspects of the models fitted in Parts I-III.

Suggested directions:
- Fit a temporal or spatial model to your own data
- Compare IID vs structured models using AIC
- Use `VP()` to partition spatial vs. environmental vs. residual variation
- Build a mixed-response `family` vector for your own data, if it has species that call for different distributions
