---
title: "Practical: extensions, spatial/temporal models and mixed responses"
subtitle: "Physalia workshop on GLLVMs"
author: "Bert van der Veen"
output: html_document
---

# Background

This practical covers two extensions to the basic GLLVM framework:

1. **Spatial and temporal autocorrelation**: accommodating non-independence of sampling units via structured covariance in the latent variables or row effects
2. **Mixed response types**: fitting models where different species have different response distributions

The key `gllvm` arguments for spatial/temporal structure are:

- `lvCor`: specifies correlation structure for the latent variables
- `row.eff`: specifies correlation structure for the row effects
- `distLV`: matrix of coordinates (or time points) for the levels used in `lvCor`'s spatial kernels -- not a precomputed distance matrix, `gllvm` computes distances internally
- `corAR1`, `corExp`, `corMatern`, `corCS`: available correlation structures

Today's lecture demonstrated these ideas on small synthetic examples, plus a real case study: the Garchinger Heide calcareous grassland data (Bauer & Albrecht 2020), combining ordinal and binomial families with a spatial correlation structure. The two component datasets from that case study (percent cover and Raunkiaer frequency, each with real GPS-derived coordinates) are available separately in the shared `data` folder; see `data/README.md` for how they relate. We use wadden and kelpforest as well:

1. wadden (abundance; used below for the mixed-response part). Its "season" covariate only has two levels (spring/summer), which is not enough to demonstrate genuine AR(1) decay: with only two time points, any correlation structure -- AR(1), compound symmetry, unstructured -- saturates the same 2x2 covariance matrix, so there is no decay over lags to show.
2. kelpforest (percent cover, `data(kelpforest)`; real site/transect/year structure spanning 21 years, 2000-2020; used below for the temporal part)
3. garchingerFrequency (Raunkiaer frequency out of 100 quadrats, real GPS coordinates for 40 plots surveyed in 1984, 1993 and 2018; used below for the spatial part)
4. Skabbholmen (counts, `data(Skabbholmen)`; real repeated transects, but only two survey years, 1978 and 1984 -- useful for a simple before/after comparison, but the same two-time-point limitation as wadden's season applies for demonstrating AR(1) decay)

# Part I: Temporal autocorrelation

We use the `kelpforest` dataset: annual surveys (2000-2020) of sessile invertebrate percent cover at 11 sites, each with up to 8 transects. Unlike a 2-level factor, 21 years is enough to actually show correlation decaying over increasing lags.

```{r setup, message=FALSE}
library(gllvm)
TMB::openmp(parallel::detectCores()-1, autopar = TRUE, DLL = "gllvm")
```

```{r kelp_setup, message=FALSE}
data(kelpforest)
Yabund <- kelpforest$Y
SPinfo <- kelpforest$SPinfo
Xenv <- kelpforest$X

# sessile invertebrates observed at least 10 times; rarer species summed into one column
Yinvert <- Yabund[SPinfo$GROUP == "INVERT"]
Yinvert10 <- Yinvert[, colSums(Yinvert > 0, na.rm = TRUE) > 9]
Yinvert10$Sum_inv <- rowSums(Yinvert[, colSums(Yinvert > 0, na.rm = TRUE) <= 9], na.rm = TRUE)

studyDesign <- data.frame(year = factor(Xenv$YEAR), site = factor(Xenv$SITE))
```

Percent cover here is bounded in [0,1) (no site ever reaches 100% cover for these species), so we use the beta-hurdle family, which handles the point mass at zero. Fit a baseline model (site-level random intercepts only) and compare to one that adds AR(1) structure across years:

```{r ar1, cache=TRUE, message=FALSE, warning=FALSE}
# Baseline: site-level random intercepts, no temporal correlation
model_iid <- gllvm(Yinvert10, family = "betaH", num.lv = 0,
                   disp.formula = rep(1, ncol(Yinvert10)),
                   studyDesign = studyDesign, row.eff = ~(1|site), n.init = 3, sd.errors = FALSE)

# AR(1) structure across years, added on top of the site-level intercepts
model_ar1 <- gllvm(Yinvert10, family = "betaH", num.lv = 0,
                   disp.formula = rep(1, ncol(Yinvert10)),
                   studyDesign = studyDesign, row.eff = ~(1|site) + corAR1(1|year), n.init = 3, sd.errors = FALSE)

AIC(model_iid, model_ar1)
```

Inspect the estimated AR(1) correlation:

```{r ar1_param}
model_ar1$params$sigma  # the ".rho"-suffixed entry is the year-to-year correlation
```

## Tasks I

1. Fit both models and compare AIC. Does adding temporal structure improve fit?
2. Interpret the AR(1) correlation parameter. How persistent is community composition from one year to the next?
3. The AR(1) correlation at lag $k$ is $\rho^k$. Using the estimated $\rho$, at what lag does the correlation drop below 0.1?
4. Try `lvCor = ~corAR1(1|year)` with `num.lv = 2` instead of putting the AR(1) structure in `row.eff`. What is the difference in interpretation?
5. Why couldn't a 2-level factor like wadden's `season` be used to demonstrate what task 3 just showed?

# Part II: Spatial models

We use `garchingerFrequency`: Raunkiaer frequency (out of 100 quadrats) for 97 species at 40 plots in the Garchinger Heide calcareous grassland, surveyed in 1984, 1993 and 2018, with real GPS-derived coordinates (see `data/README.md`).

```{r spatial_setup, message=FALSE}
Yf <- read.csv("../../data/garchingerFrequencyY.csv", row.names = 1)
Xf <- read.csv("../../data/garchingerFrequencyX.csv")

plot_f <- factor(Xf$PlotID)
# Plots repeat across the 3 survey years (120 rows, 40 distinct plots).
# distLV needs one coordinate row per distinct level of the lvCor grouping
# variable, not one row per observation.
coords_plot <- as.matrix(Xf[match(levels(plot_f), Xf$PlotID), c("E", "N")])
```

Fit a baseline model (no spatial structure) and compare to one with exponentially decaying correlation between plots:

```{r spatial, cache=TRUE, message=FALSE, warning=FALSE}
model_iid <- gllvm(Yf, family = "binomial", Ntrials = 100, num.lv = 2,
                   n.init = 5, seed = 1:5, sd.errors = FALSE)

model_spatial <- gllvm(Yf, family = "binomial", Ntrials = 100, num.lv = 2,
                       lvCor = ~corExp(1|plot),
                       studyDesign = data.frame(plot = plot_f),
                       distLV = coords_plot, n.init = 5, seed = 1:5, sd.errors = FALSE)

AIC(model_iid, model_spatial)
```

`corExp` model fits are sensitive to starting values: with too few restarts, one of the two LV ranges can collapse to a degenerate, effectively-zero value. `n.init = 5` with fixed seeds avoids that here; if you still see a `rho.lv` value near zero (e.g. `1e-70`), increase `n.init` further.

Inspect the estimated spatial range:

```{r spatial_param}
model_spatial$params$rho.lv  # range per LV, in the same units as the coordinates (km)
```

The range is the distance at which correlation decays to $e^{-1}\approx0.37$; the effective range (correlation below 0.05) is about $3\times$ that.

## Tasks II

1. Fit both models and compare AIC. A lower log-likelihood for the spatial model does not automatically mean a worse model: `gllvm`'s variational approximation bound is not equally tight for every `lvCor` structure, so logLik comparisons across different correlation structures should be read cautiously.
2. Interpret the estimated range parameter(s). What is the effective range, and how does that compare to the extent of the study area (all plots fall within roughly 300m x 300m)?
3. Use `VP()` to see how much variation the spatial structure explains.
4. Why would giving `distLV` one coordinate row per observation (120 rows) instead of one per distinct plot (40 rows) be wrong here, even though every row does have its own coordinates?
5. If you have your own spatial dataset, try fitting a `corExp` model to it. For large datasets, consider a sparse approximation (`Lambda.struc = "bdNN"` or `"UNN"` with `NN` set to a small number of neighbours).

# Part III: Mixed response types

Different species in a single dataset can call for different response distributions — for example, rare species recorded as presence-absence and common species recorded as counts. As covered in the lecture, `gllvm` supports this by passing a **character vector** to `family`, with one entry per species column (matching column order), rather than a single distribution for the whole community.

We use the wadden data again. To illustrate, we will treat the first 10 species as presence-absence (thresholding to 0/1) and leave the rest as counts:

```{r mixed_setup, message=FALSE}
Y <- read.csv("../../data/waddenY.csv")[, -c(1:2)]
Y <- Y[, colSums(ifelse(Y == 0, 0, 1)) > 2]
```

```{r mixed, cache=TRUE, message=FALSE, warning=FALSE}
Y_mixed <- Y
Y_mixed[, 1:10] <- ifelse(Y_mixed[, 1:10] > 0, 1, 0)

family_vec <- c(rep("binomial", 10), rep("negative.binomial", ncol(Y_mixed) - 10))

model_mixed <- gllvm(Y_mixed, num.lv = 2, family = family_vec, n.init = 3, sd.errors = FALSE)
```

## Tasks III

1. Compare `model_mixed` to a model that (incorrectly) treats all species in `Y_mixed` as counts (`family = "negative.binomial"`). What goes wrong for the thresholded species?
2. Think about your own data: do different species call for different distributions? Build a `family` vector for it, matching column order.
3. Check `?gllvm` for the full list of families that can be mixed together in one model.
