---
title: "Practical: predicting species richness and diversity"
subtitle: "Physalia workshop on GLLVMs"
author: "Bert van der Veen"
output: html_document
---

# Background

In this practical we use a multispecies model to predict species richness and explore diversity components. The key ideas from the lecture are:

- Species richness is an emergent property of the joint species model
- The correct distribution for species richness is the Poisson-Binomial
- `predictSR()` computes expected richness from any fitted `gllvm` model
- Variance components connect to alpha, beta, and gamma diversity

We will use the alpine plant dataset (binary presence-absence) which we have used before. The goal is to fit a multispecies model, predict species richness along the elevation gradient, and explore how model components relate to diversity.

```{r setup, message=FALSE}
library(gllvm)
TMB::openmp(parallel::detectCores()-1, autopar = TRUE, DLL = "gllvm")

Y <- read.csv("../../data/alpineY.csv")[,-1]
X <- read.csv("../../data/alpineX.csv")[,-1]
X <- X[rowSums(Y) > 0, ]
Y <- Y[rowSums(Y) > 0, ]
X_sc <- data.frame(scale(X[, sapply(X, is.numeric)]))
```

# Part I: Fitting a model and predicting richness

We start by fitting a model with elevation as a covariate and two latent variables.

```{r fit, cache=TRUE, message=FALSE}
model1 <- gllvm(y = Y, X = X_sc,
                formula = ~ELEV,
                num.lv = 2,
                family = "binomial",
                sd.errors = FALSE,
                optim.method = "L-BFGS-B",
                seed = 1)
```

Now predict species richness along the elevation gradient:

```{r predict}
elev_seq <- seq(min(X_sc$ELEV), max(X_sc$ELEV), length.out = 100)
newX <- data.frame(ELEV = elev_seq)

sr <- predictSR(model1, newX = newX, level = 0)
```

Plot observed richness against predicted richness:

```{r plot_sr}
obs_sr <- rowSums(Y)
plot(X_sc$ELEV, obs_sr, pch = 16, col = rgb(0,0,0,0.3),
     xlab = "Elevation (scaled)", ylab = "Species richness")
lines(elev_seq, sr$expected$fit, col = "red", lwd = 2)
lines(elev_seq, sr$expected$lower, col = "red", lwd = 1, lty = 2)
lines(elev_seq, sr$expected$upper, col = "red", lwd = 1, lty = 2)
```

Check residuals from the Poisson-Binomial distribution:

```{r resid}
plot(residuals(sr))
```

## Tasks I

1. Fit the model and predict species richness as above. Does the model capture the observed richness pattern well?
2. Add more covariates (e.g., `SLOPE`, `MIND`, `SOLRAD`) and see if the richness prediction improves.
3. Change the number of latent variables (0, 1, 2, 3). How does this affect the richness prediction?
4. Use `AIC()` or `BIC()` to compare models. Does including latent variables improve fit?
5. Examine the `predictSR` residuals: do they look like random noise, or is there remaining structure?

# Part II: Diversity components

The variance components from the model connect to ecological diversity concepts.

```{r var_components}
# Variation explained by the latent variables
getResidualCov(model1)$trace

# Relative variation per axis
var.q <- getResidualCov(model1)$var.q
var.q / sum(var.q)
```

For the species-specific elevation effects (treated as random), the variance tells us how much species differ in their response to elevation, a measure of niche differentiation driving beta diversity.

```{r beta_diversity, eval=FALSE}
model2 <- gllvm(y = Y, X = X_sc,
                formula = ~(0 + ELEV|1),
                num.lv = 0,
                family = "binomial",
                sd.errors = FALSE)
summary(model2)  # examine variance of elevation slopes
```

## Tasks II

1. Extract and interpret the residual covariance trace: how much variation is captured by the latent variables?
2. Compare the relative variation explained by each latent variable axis.
3. Fit a model with a random slope for elevation across species. What is the variance of species' elevation responses?
4. Try predicting species richness with the random-slope model. How does the uncertainty compare to the fixed-effect model?
5. Think about what alpha, beta, and gamma diversity mean in the context of this dataset and model. Which model components drive each component of diversity?

# Part III: Exploring the Poisson-Binomial

To understand why univariate species richness analysis can be misleading, compare the predicted distribution of species richness from your model to what a Poisson regression would give.

```{r pb_vs_pois}
# Expected richness and variance from the model
pred_probs <- predict(model1, type = "response") # n x m matrix
exp_sr <- rowSums(pred_probs)        # Poisson-Binomial mean
var_sr <- rowSums(pred_probs * (1 - pred_probs))  # PB variance

# Compare to Poisson assumption
plot(exp_sr, var_sr, xlab = "Expected species richness",
     ylab = "Predicted variance", pch = 16, col = "steelblue")
abline(0, 1, col = "red", lty = 2, lwd = 2)  # Poisson: var = mean
legend("topleft", legend = c("Poisson-Binomial", "Poisson (var=mean)"),
       pch = c(16, NA), lty = c(NA, 2), col = c("steelblue", "red"))
```

## Tasks III

1. Run the code above. Is the Poisson assumption (variance = mean) realistic for these data?
2. When variance < mean (underdispersion), what does that imply ecologically?
3. Subset the data to a low-diversity region (e.g., high elevation). How does the Poisson approximation perform there versus a diverse region?
