---
title: "Practical: extensions — spatial/temporal models and mixed responses"
subtitle: "Physalia workshop on GLLVMs"
author: "Bert van der Veen"
output: html_document
---

# Background

This practical covers two extensions to the basic GLLVM framework:

1. **Spatial and temporal autocorrelation** — accommodating non-independence of sampling units via structured covariance in the latent variables or row effects
2. **Mixed response types** — fitting models where different species have different response distributions *(placeholder — see note below)*

The key `gllvm` arguments for spatial/temporal structure are:

- `lvCor` — specifies correlation structure for the latent variables
- `row.eff` — specifies correlation structure for the row effects
- `distLV` — matrix of distances between sites (for spatial models)
- `corAR1`, `corExp`, `corMatern`, `corSym` — available correlation structures

# Part I: Temporal autocorrelation

We use a subset of the wadden sea data which has a temporal structure (seasons).

```{r setup, message=FALSE}
library(gllvm)
TMB::openmp(parallel::detectCores()-1, autopar = TRUE, DLL = "gllvm")

Y <- read.csv("../../data/waddenY2.csv")[, -c(1:2)]
Y <- Y[, colSums(ifelse(Y == 0, 0, 1)) > 2]
X <- read.csv("../../data/waddenX.csv")
X <- X[, !apply(X, 2, anyNA)]
X[, unlist(lapply(X, is.numeric))] <- scale(X[, unlist(lapply(X, is.numeric))])
```

Fit a baseline model (IID latent variables) and compare to one with AR(1) structure across seasons:

```{r ar1, cache=TRUE, message=FALSE, warning=FALSE}
# Baseline: independent latent variables
model_iid <- gllvm(Y, num.lv = 2, family = "tweedie", Power = NULL,
                   disp.formula = rep(1, ncol(Y)), n.init = 3)

# AR(1) structure across seasons in the row effect
model_ar1 <- gllvm(Y, num.lv = 2, family = "tweedie", Power = NULL,
                   disp.formula = rep(1, ncol(Y)),
                   studyDesign = X[, "season", drop = FALSE],
                   row.eff = ~corAR1(1|season), n.init = 3)

AIC(model_iid, model_ar1)
```

Inspect the estimated AR(1) correlation:

```{r ar1_param}
model_ar1$params$rho.row  # temporal correlation between seasons
```

## Tasks I

1. Fit both models and compare AIC. Does adding temporal structure improve fit?
2. Interpret the AR(1) correlation parameter. Is community composition persistent across seasons?
3. Try `lvCor = ~corAR1(1|season)` instead of `row.eff`. What is the difference in interpretation?
4. Visualize the ordination from both models. Does temporal structure change the ordination?

# Part II: Spatial models

If you have site coordinates, you can fit a model with spatial correlation. Here we simulate coordinates for illustration — in your own data, use real coordinates.

```{r spatial_setup}
# Simulate spatial coordinates for the wadden sites
set.seed(42)
n_sites <- nrow(Y)
coords <- data.frame(
  lon = runif(n_sites, 0, 100),
  lat = runif(n_sites, 0, 100)
)
dist_mat <- as.matrix(dist(coords))
```

```{r spatial, cache=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
model_spatial <- gllvm(Y, num.lv = 2, family = "tweedie", Power = NULL,
                       disp.formula = rep(1, ncol(Y)),
                       lvCor = ~corExp(1|site),
                       studyDesign = data.frame(site = factor(1:nrow(Y))),
                       distLV = dist_mat,
                       Lambda.struc = "UNN", NN = 5,
                       n.init = 1)
```

Note: spatial models are computationally demanding. With real data, ensure you have sufficient memory and consider sparse approximations (`Lambda.struc = "bdNN"` or `"UNN"` with `NN` set to a small number of neighbours).

## Tasks II

1. Read through the code for fitting a spatial model. What does `distLV` do?
2. What does the `NN` argument control? What are the trade-offs of using more vs. fewer neighbours?
3. How would you use `VP()` to see how much variation the spatial structure explains?
4. If you have your own spatial dataset, try fitting a `corExp` model to it.

# Part III: Mixed response types *(placeholder)*

Mixed response types — where different species have different distributions within a single joint model — are a newer feature in `gllvm`. For example, some species might be recorded as binary presence-absence while others are recorded as counts.

```{r mixed_placeholder, eval=FALSE}
# Example structure (functionality in active development)
# model_mixed <- gllvm(y = Y,
#                      family = list("binomial", "poisson", "negative.binomial"),
#                      ...)
```

Check the `gllvm` package documentation and vignettes for the current state of mixed response type support:

```{r help, eval=FALSE}
?gllvm
vignette(package = "gllvm")
```

## Tasks III

1. Check the `gllvm` help page and package vignettes for mixed response types. What families are currently supported together?
2. Think about your own data: do different species in your dataset call for different distributions? What combinations would you need?

# Part IV: Own data

Use the remaining time to apply what you have learned to your own dataset, or to explore further aspects of the models fitted in Parts I-III.

Suggested directions:
- Fit a temporal or spatial model to your own data
- Compare IID vs structured models using AIC
- Use `VP()` to partition spatial vs. environmental vs. residual variation
- Explore the `gllvm` documentation for the latest features on mixed responses
